Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-21 paper2.qmd
section:4; process:write

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates  [@Perron_et_al_2015, pp. 2]. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify statistical modeling, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the apparent disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.

<!-- 4. what is the organization of the study? -->
To address these issues, the study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods. As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. @sec-discussion discusses the findings, limitations, and challenges and explores avenues for future research. Finally, @sec-conclusion summarizes the study's conclusions.

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Despite its limitations, Case V remains the most widely used model in CJ literature. This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case. The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). However, as @vanderLinden_et_al_2017_I [pp. 16] noted, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation. Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

@@

<!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
Building on the previous section, it is clear that, despite its limitations, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along such trait continuum [@Thurstone_1927b, pp. 269; @Everitt_et_al_2010]. Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests. For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty. Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. Notably, the direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. @fig-measurement_error illustrates these issues, showing how neglecting the uncertainty associated with BTL scores can reduce the precision of an effect linked to the predictor.

@@

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp. 4]. *Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010]. 
<!-- The causal assumptions encoded in a DAG constitute the theoretical model on which identification analysis rests. Identification analysis is concerned with determining whether or not the true value of a parameter can be estimated using the assumptions in the model (graph), regardless of random variability in the data due to small samples [@Schuessler_et_al_2023, pp. 4]. -->

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear throughout the document clarifies these concepts. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, researcher design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli), to produce a dichotomous outcome that reflects the relative quality of each text (trait). Based on this setup, researchers reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Finally, following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 6. What the section will do? -->
Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone’s general form using the structural approach to causal inference. Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition to enhancing statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.

<!-- 7. How the section is divided? -->
Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we call the *conceptual-population model*. This model assumes that researchers have access to a *census* of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population, hence its name. Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we call the *sample-comparison model*. This model assumes a more realistic scenario where researchers only have access to a sample of judges, individuals, stimuli, and comparisons from the conceptual population.

@@

### Integrating initial theoretical principles {#sec-theory-theoretical_P1}

@@

<!-- 2. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] presents the initial integration of theoretical principles into the conceptual-population model for the example, applied to examine the effect of different teaching methods on students' writing ability. This SCM maps the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables. The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli). Moreover, the index $j$ refers to the judge, and $k$ specifies the judgment number, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp. 366-376]. Notably, the indexing system allows comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible). However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical meaning within the CJ framework.

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory). Additionally, based on the arguments developed under the first scenario of @sec-theory-issue1b,  and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$. Together with the outcome, these variables form the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$. Finally, the SCM shows the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

<!-- 4. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a graph consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence, while the absence of an edge implies no relationship between the nodes. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs conventionally represent observed variables with solid black circles and unobserved (latent) variables with open circles [@Morgan_et_al_2014]. They also display conditioned variables by surrounding them with a square. Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp. 32]. While DAGs often omit exogenous variables for simplicity, following a *standard representation*, this section adopts the *magnified representation* of DAGs, as including exogenous variables can improve clarity and help reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- 5. the preliminary DAG for CJ -->
@fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the causal relationships among variables based on Thurstone's theory. Particularly, the graph shows how the texts' discriminal processes $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$. It also depicts the judges' biases $(B_{kj})$ as an additional factor influencing the texts' discriminal difference. Furthermore, the DAG distinguish between observed endogenous variables, such as the outcome, and latent endogenous variables, which include the texts' discriminal processes, their discriminal difference, and the judges' biases.

### The comparative census data structure {#sec-theory-theoretical_P2}

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system, outlined in the previous section, into an simpler equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

@@

<!-- 2. what is the data structure? -->
where an element of $I$ corresponds to a unique individual $h$ or $i$, while each row of $IA$ corresponds to a unique pairing of individuals $h, i$ with stimuli $a, b$. Here, $n_{I}$ represents the number of individuals, while $n_{A}$ denotes the number of stimuli available per individual. Thus, vector $I$ has a length of $n_{I}$, and matrix $IA$ has a dimension of $(n_{I} \cdot n_{A})$ rows and $2$ columns. The vector and the matrix consist of scalars from the set of natural numbers $\mathbb{N}$.

Similarly, an element of the vector $J$ corresponds to a unique judge $j$. Moreover, each row of $JK$ specifies a unique pairing of a judge $j$ with a judgment number $k$. In this case, $n_{J}$ indicates the total number of judges, and $n_{K}$ represents the number of judgments each judge performs. As a result, vector $J$ has a length of $n_{J}$, and matrix $JK$ has a dimension of $(n_{J} \cdot n_{K})$ rows and $2$ columns. Like the previous structures, the vector and the matrix consist of scalars from the set of natural numbers $\mathbb{N}$.

Additionally, we define the matrix $RPUV$, where each row uniquely pairs the corresponding rows from the matrices $IA$ and $JK$. This matrix reflects the fundamental idea that the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions. Thus, the matrix is defined as follows:

@@

This matrix also consists of scalars from the set of natural numbers $\mathbb{N}$ and has a dimension of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns. The matrix's dimension results from the conceptual-population model's assumption that researchers have access to a *census* of comparative data, capturing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of individual-stimuli comparison pairs within the conceptual population. The coefficient is defined as follows:

$$
{n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! }
$$

We visualize the structure of these vectors and matrices by considering small values for the parameters corresponding to number of individuals, the number of available stimuli per individual, the total number of judges, and the number of judgments each judge performs. For instance, assuming $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$, the matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:

@@

<!-- 3. how de we use that data structure? -->
Thus, using the data structure defined by the matrices in equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in a simpler equivalent vectorized form, as shown in @fig-cj04. In this depiction, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values. These vectors encompass all values from the conceptual population. Specifically, $O^{cp}_{RPUV}$ is an observed vector of length ${n{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, while $D_{RPUV}$ is a latent vector of the same dimension. Additionally, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.

@@

<!-- 1. What are we going to do? -->
Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model incorporates two *hierarchical structural components*. Each component describes how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010]. This addition addresses the problems arising when an additional hierarchical structure is relevant to the stimuli, as outlined in the second scenario of @sec-theory-issue1b. Furthermore, it facilitates analyses that integrate both trait measurement and hypothesis testing processes, thereby accounting for both traits and their associated uncertainties, reducing bias and improving measurement precision (see @sec-theory-issue2).

<!-- principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and IRT [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.  -->

<!-- 2 Explaining what we did -->
@fig-cj09 displays the two hierarchical structural components for the example, applied to examine the effect of different teaching methods on students' writing ability. Specifically, the top branch of [DAG -@fig-cj09_dag] depicts the first component. This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$, such as text length, causally influence the texts' written-quality trait $(T_{IA})$. The DAG also highlights the idiosyncratic errors $(e_{IA})$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain. Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively. Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns.

@@

<!-- 3. segway for next section -->
Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issues. These enhancements include accounting for judges' biases and integrating hierarchical structural components. Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *census* of comparative data, encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population. Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario. The extensions to the conceptual-population model, derived from this realistic scenario, are concisely described in the sample-comparison model.

>>>>

<!-- 3. But what critical issues remain in CJ Research? -->
Nevertheless, despite the increasing number of CJ studies, research in this domain remains unsystematic and fragmented, leaving several critical issues unresolved. This study identifies and discusses two prominent issues of traditional approaches that can undermine the reliability and validity of CJ's trait estimates  [@Perron_et_al_2015, pp. 2]. First, it critiques the heavy reliance on Thurstone's Case V assumptions [@Thurstone_1927b] and, by extension, the Bradley-Terry-Luce (BTL) model [@Bradley_et_al_1952; @Luce_1959] when analyzing CJ data. Specifically, the study raises concerns about the assumptions of equal discriminal dispersions and zero correlation between the stimuli. While these assumptions simplify statistical modeling, they may fail to capture the complexity of CJ data, potentially leading to unreliable and inaccurate trait estimates. Second, the study highlights the disconnect between CJ's trait measurement and hypothesis testing processes. Although separating these processes simplifies the analysis of CJ data, it may also undermine the reliability of various statistical results derived from the measurement.

<!-- 5. So, what are we doing? -->
To address these issues, this study extends Thurstone's general form using a systematic and integrated approach based on Causal and Bayesian inference methods. In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the BTL model assumptions to actively testing whether those assumptions fit the data under analysis.

<!-- Many cross-cultural scientists already pay close attention to concerns of causal inference and comparison without use of a formal framework (e.g., Norenzayan & Heine, 2005; Pollet et al., 2014). For these researchers, a formalized framework can provide a vocabulary to articulate their concerns and work toward solutions in a more systematic manner. [@Deffner_et_al_2022, pp. 2] -->

<!-- Specifically, a framework is needed which lets us reason about the causes of non-invariance. [@Sterner_et_al_2024, pp. 751] -->

<!-- 4. what is the organization of the study? -->
As a result, the study divides its content into six main sections. @sec-thurstone_theory provides an overview of Thurstone's theory. @sec-theory-issues discusses the identified issues in detail. @sec-theoretical extends Thurstone's general form to address these challenges. The extension integrates core theoretical principles alongside key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. @sec-statistical translates these theoretical and practical elements into a probabilistic statistical model to analyze dichotomous pairwise comparison data. @sec-discussion discusses the findings, limitations, and challenges and explores avenues for future research. Finally, @sec-conclusion summarizes the study's conclusions.

@@

<!-- 1. The Case V and its simpler statistical representation: the BTL model -->
Despite its limitations, Case V remains the most widely used model in CJ literature. This preference primarily stems from the BTL model, which provides a simplified statistical representation of the case. The BTL model mirrors the assumptions of Case V, with one notable distinction: while Case V assumes a Normal distribution for the stimuli' discriminal processes, the BTL model uses the more mathematically tractable Logistic distribution [@Andrich_1978; @Bramley_2008, pp. 254] (see @tbl-thurstone_cases). However, this substitution has minimal impact on the model's estimation or interpretation because the discriminal process scale is arbitrary up to a non-monotonic transformation [@vanderLinden_et_al_2017_I, pp. 16; @McElreath_2022]. Furthermore, this limited impact is supported by the fact that the Normal and Logistic distributions exhibit analogous statistical properties, differing only by a scaling factor of approximately $1.7$.

@@

<!-- 1. Utilizing BTL Scores for further analysis and hypothesis testing -->
Building on the previous section, it is clear that, despite its limitations, researchers typically rely on the BTL model to measure a trait and place the compared stimuli along its continuum [@Thurstone_1927b, pp. 269]. Additionally, the CJ literature shows that researchers frequently use point estimates of BTL scores or their transformations to conduct further analyses or hypothesis tests. For example, researchers have used these scores to identify 'misfit' judges and stimuli [@Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018], detect biases in judges' ratings [@Pollitt_et_al_2003; @Pollitt_2012b], calculate correlations with other assessment methods [@Goossens_et_al_2018; @Bouwer_et_al_2023], or test hypotheses related to the underlying trait of interest [@Casalicchio_et_al_2015; @Bramley_et_al_2019; @Boonen_et_al_2020; @Bouwer_et_al_2023; @vanDaal_et_al_2017; @Jones_et_al_2019; @Gijsen_et_al_2021].

<!-- 3. But what are the challenges in using BTL scores for further analysis? -->
Nevertheless, while separating the trait measurement and hypothesis testing processes simplifies the analysis of CJ data, the statistical literature cautions against relying solely on the point estimates of BTL scores to conduct further analyses or hypothesis tests, as this practice can undermine the resulting statistical conclusions. A key consideration is that BTL scores are parameter estimates that inherently carry uncertainty (measurement error). Ignoring this uncertainty can bias the analysis and reduce the precision of hypothesis tests. The direction and magnitude of such biases are often unpredictable. Results may be attenuated, exaggerated, or remain unaffected depending on the degree of uncertainty in the scores and the actual effects being tested [@Kline_et_al_2023, pp. 25; @Hoyle_et_al_2023, pp. 137]. Furthermore, the reduced precision in hypothesis tests diminishes their statistical power, increasing the likelihood of committing type-I or type-II errors [@McElreath_2020]. @fig-measurement_error illustrates these issues, showing how neglecting measurement error in an outcome can reduce the precision of a predictor' estimated effect.

@@

<!-- 1. The structural approach to causal inference -->
The *structural approach* to causal inference provides a formal framework for identifying causes and estimating their effects using data. The approach uses structural causal models (SCMs) and directed acyclic graphs (DAGs) [@Pearl_2009; @Pearl_et_al_2016; @Gross_et_al_2018; @Neal_2020] to formally and graphically represent the assumed causal structure of a system, such as the one found in CJ experiments. Essentially, SCMs and DAGs function as *conceptual models* on which identification analysis rests [@Schuessler_et_al_2023, pp. 4]. *Identification analysis* helps researchers to determine whether an estimator can accurately compute an estimand based solely on its (causal) assumptions, regardless of random variability [@Schuessler_et_al_2023, pp. 4]. Here, *estimands* represent the specific quantities researchers aim to determine [@Everitt_et_al_2010]. *Estimators* denote the methods or functions that transform data into an estimate, while *estimates* are the numerical values approximating the estimand [@Neal_2020; @Everitt_et_al_2010].

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example that will appear throughout the document clarifies these concepts. In this example, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, a researcher designs a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students' written texts (stimuli) to produce a dichotomous outcome reflecting the relative quality of each text (trait). Based on this setup, researchers can reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 6. What the section will do? -->
Thus, this section addresses the issues identified in @sec-theory-issues by extending Thurstone's general form using the structural approach to Causal inference. Specifically, it combines the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons. In addition to improving statistical accuracy and strengthening measurement reliability and validity, the approach offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing whether those assumptions fit the data under analysis.

<!-- 7. How the section is divided? -->
Accordingly, @sec-theory-theoretical_P incorporates the theoretical principles into what we refer to as the *conceptual-population model*. This model assumes that researchers have access to a "conceptual population" of comparative data, that is, data representing all repeated judgments made by every available judge for each pair of stimuli produced by each pair of individuals in the population. Conversely, @sec-theory-theoretical_SC integrates the assessment design features into what we refer to as the *sample-comparison model*. This model assumes a more realistic scenario where researchers only have access to a sample of judges, individuals, stimuli, and comparisons from the conceptual population.

@@

@@

<!-- 2. The DAGs -->
Notably, every SCM has an associated DAG [@Pearl_et_al_2016; @Cinelli_et_al_2020]. A DAG is a *graph* consisting of nodes connected by edges, where nodes represent random variables. The term *directed* indicates that edges or arrows extend from one node to another, indicating the direction of causal influence. The absence of an edge implies no direct relationship between the nodes. The term *acyclic* means that the causal influences do not form loops, ensuring the influences do not cycle back on themselves [@McElreath_2020]. DAGs conventionally depict observed variables as solid black circles and unobserved (latent) variables as open circles [@Morgan_et_al_2014]. They also display conditioned variables by surrounding them with a square. Here, *conditioning* refers to restricting the statistical analysis to a subset of the population defined by the values of the given variable [@Neal_2020, pp. 32]. Although DAGs often omit exogenous variables for simplicity, this section includes exogenous variables to improve clarity and reveal potential issues related to conditioning and confounding [@Cinelli_et_al_2020].

<!-- the *standard representation* omit exogenous variables for simplicity -->
<!-- the *magnified representation* includes exogenous variables -->

### Integrating the first theoretical principles {#sec-theory-theoretical_P1}

<!-- 1. the preliminary SCM for CJ -->
[SCM -@fig-cj03_scm] illustrates the first theoretical principles of the conceptual-population model designed to examine the impact of different teaching methods on students' writing ability. This SCM outlines the relationship between the conceptual-population outcome $(O^{cp}_{hiabjk})$ and several related variables. The subscripts $h$ and $i$ label the students who authored the texts (individuals), while the indices $a$ and $b$ represent the compared texts (stimuli). The index $j$ refers to the judge, and $k$ represents the judgment index, accounting for experimental conditions where judges perform the same judgment multiple times, i.e., a *repeated measures design* [@Lawson_2015, pp. 366-376]. Notably, the indexing system facilitates comparisons between texts written by the same student ($h = i$, $a \neq b$) and between texts written by distinct students ($h \neq i$, where $a = b$ is possible). However, it excludes cases where judges compare a student's text to itself ($h = i$, $a = b$), as such comparison lacks practical relevance within the CJ framework.

In line with Thurstone's theory, [SCM -@fig-cj03_scm] depicts the texts' discriminal processes $(T_{ha}, T_{ib})$ and their discriminal difference $(D_{hiabjk})$ (see @sec-thurstone_theory). Additionally, based on the arguments developed in @sec-theory-issue1b and the recommendations of @Andrich_1978 and @Wainer_et_al_1978, the SCM incorporates the judges' biases $(B_{kj})$. Together with the outcome, these variables constitute the preliminary set of endogenous variables, $V = \{ O_{hiabjk}, D_{hiabjk}, T_{ha}, T_{ib}, B_{kj} \}$. Finally, the SCM presents the preliminary set of structural equations, $F = \{ f_{O}, f_{D} \}$, which define the non-parametric dependencies among these variables.

@@

<!-- 2. the preliminary DAG for CJ -->
Moreover, @fig-cj03_dag displays the DAG corresponding to [SCM -@fig-cj03_scm], illustrating the expected causal outlined in Thurstone's theory. The graph shows that the discriminal processes of the texts $(T_{ha}, T_{ib})$ influence their discriminal difference $(D_{hiabjk})$, which in turn determines the outcome $(O^{cp}_{hiabjk})$. It also highlights the influence of judges' biases $(B_{kj})$ on the discriminal difference. Additionally, the DAG differentiates between observed endogenous variables, such as the outcome, and latent endogenous variables, including the texts' discriminal processes, their discriminal difference, and the judges' biases.

<!-- To use graphs to reason about confounding control and other forms of bias requires that the representation on the graph be such that there is: -->
<!-- 1. No ungraphed confounding: Any variable that affects two or more variables in the graph via distinct direct effects must be shown on the graph as an ancestor of the affected variables, even if it is unmeasured. -->
<!-- 2. No ungraphed collider stratification: Any variable that affects selection or is used for stratification and is affected by two or more variables in the graph (including its random disturbance) must be shown on the graph as a descendant of the variables affecting it, even if it is unmeasured. -->
<!-- As may become clear from the descriptions below, these requirements are needed to ensure that any association between two variables in the graph can be atributed solely to open paths between the variables within the graph, and thus can be traced to either variation in graphed causal variables, or conditioning on effects of such variables, or both. Conditions 1 and 2 enforce the connectivity requirement of DAGs by demanding the graph we draw show all variables which can causally explain associations (either as uncontrolled confounders or as controlled colliders), even if those variables are unmeasured or hypothetical. [@Lash_et_al_2021, pp. 113] -->

### The conceptual-population data structure {#sec-theory-theoretical_P2}

<!-- 1. Why is useful to declare the data structure? -->
Although specifying a system's data structure is not mandatory when using SCMs and DAGs, defining one can improve clarity and facilitate the description of the system. Thus, to re-express the scalar form of the CJ system outlined in the previous section into an equivalent vectorized form, we first define the vectors $I$ and $J$, along with the matrices $IA$ and $JK$, as follows:

@@

<!-- 2. what is the data structure? -->
Here, each element of $I$ represents a unique individual $h$ or $i$, where $n_{I}$ denotes the total number of individuals. Similarly, each element of $J$ corresponds to a unique judge $j$, with $n_{J}$ indicating the total number of judges. Moreover, each row of $IA$ represents a unique pairing of individuals $h, i$ with stimuli $a, b$. As a result, the matrix $IA$ contains $(n_{I} \cdot n_{A})$ rows and $2$ columns, where $n_{A}$ specifies the number of stimuli available per individual. Likewise, each row of $JK$ associates a judge $j$ with a judgment index $k$, where $n_{K}$ indicates the number of judgments each judge makes. Consequently, the matrix $JK$ has $(n_{J} \cdot n_{K})$ rows and $2$ columns.

Additionally, we define the matrix $RPUV$ that assigns each row from the $IA$ matrix to a corresponding row from the $JK$ matrix. Thus, the matrix consists of $n = {n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$ rows and $6$ columns. Here, the term ${n_{I} \cdot n_{A} \choose 2}$ represents the Binomial coefficient, which quantifies the total number of repeated judgments each judge makes for every pair of stimuli generated by each pair of individuals in the population. The matrix is defined as follows:

<!-- $$ -->
<!-- {n_{I} \cdot n_{A} \choose 2} = \frac{ (n_{I} \cdot n_{A})! }{ 2! (n_{I} \cdot n_{A} - 2)! } -->
<!-- $$ -->

@@

It is easier to visualize the structure of these vectors and matrices by considering an example where $n_{I} = 5$, $n_{A} = 2$, $n_{J} = 3$, and $n_{K} = 3$. In this simple case, the vectors and matrices described in equations ([-@eq-mat12]) and ([-@eq-mat3]) take the following form:

@@

<!-- 3. how de we use that data structure? -->
Then, using equations ([-@eq-mat12]) and ([-@eq-mat3]), we can re-express [SCM -@fig-cj03_scm] and [DAG -@fig-cj03_dag] in an equivalent vectorized form, as shown in @fig-cj04. In this depiction, the outcome $O^{cp}_{RPUV}$, the texts' discriminal difference $D_{RPUV}$, their discriminal processes $T_{IA}$, and the judges' biases $B_{JK}$ are represented as vectors rather than scalar values. These vectors capture all the observations from the conceptual population. Specifically, $O^{cp}_{RPUV}$ is an observed vector of length $n={n_{I} \cdot n_{A} \choose 2} \cdot n_{J} \cdot n_{K}$, and $D_{RPUV}$ is a latent vector of the same dimension. Moreover, $T_{IA}$ and $B_{JK}$ are latent vectors of lengths $(n_{I} \cdot n_{A})$ and $(n_{J} \cdot n_{K})$, respectively.

@@

<!-- 1. What are we going to do? -->
Building on the principles of Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24], the conceptual-population model incorporates two *hierarchical structural components*. A structural component describes how different observed or latent variables influence the primary latent variable of interest [@Everitt_et_al_2010]. Moreover, their hierarchical nature enable analyses when the stimuli exhibit a hierarchical structure (refer to @sec-theory-issue1b) and when researchers aim to test hypotheses while accounting for the uncertainties associated with the traits (see @sec-theory-issue2).

<!-- 2 Explaining what we did -->
@fig-cj09 displays these components applied to examine the effect of different teaching methods on students' writing ability. The top branch of [DAG -@fig-cj09_dag] depicts the first component. This component describes how the students' writing-quality trait $(T_{I})$ and a set of *relevant* text-related variables $(X_{IA})$, such as text length, causally influence the texts' written-quality trait $(T_{IA})$. The DAG also highlights the idiosyncratic errors $(e_{IA})$, which capture variations in the texts' trait that neither the students' trait nor other relevant text-related variables can explain. Here, $T_{I}$ and $e_{IA}$ are latent vectors of length $n_{I}$ and $(n_{I} \cdot n_{A})$, respectively. Additionally, $X_{IA}$ is an observed matrix with a dimension of $(n_{I} \cdot n_{A})$ rows and $q_{IA}$ columns.

@@

<!-- 3. segway for next section -->
Overall, the conceptual-population model extends Thurstone's general form by introducing key innovations to address the limitations discussed in @sec-theory-issues. These enhancements include accounting for judges' biases and integrating hierarchical structural components. Nevertheless, despite its promise of enhancing measurement accuracy and precision, the model still depends on the unrealistic assumption that researchers have access to a *conceptual population* of comparative data, encompassing all repeated judgments from every judge for each pair of stimuli produced by every individual in the population. Since researchers rarely meet this assumption in practice, they must consider a more realistic scenario. The extensions to the conceptual-population model, derived from this realistic scenario, are concisely described in the sample-comparison model.
